---
output: html_document
editor_options: 
  chunk_output_type: console
---
Exploratory Data Analysis of White Wine Quality by ADENIYI EZEKIEL AYOTOMIWA
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(dplyr)
library(reshape2)
theme_set(theme_minimal())
```

```{r echo=FALSE, Load_the_Data}

# Load the Data
getwd()
setwd('C:/Users/SURFACE/Desktop/Final Project')
  
#load the white wine quality dataset

wwq <- read.csv('wineQualityWhites.csv')
#inspecting the data set

names(wwq)

#structure of dataset
str(wwq)

#converting to a factor
wwq$X <- factor(wwq$X)
str(wwq$X)
#converting to a factor
wwq$quality <- factor(wwq$quality)

str(wwq$quality)
# there are 4898 observations made in our dataset and there are 11
#variables. Variable x and quality are an integer number.



```

# Univariate Plots Section
Taking Plots of each of the variables.
```{r echo=FALSE, Univariate_Plots}

#names of our dataset
names(wwq)

#plotting the fixed acidity, binwidth of 0.1
qplot(fixed.acidity, 
      data = wwq,
      binwidth = 0.1,  
      xlim = c(4, 10),
      xlab = 'fixed acidity',
      color = I('black'),
      fill = I('pink'))
#median and mean is very close.
summary(wwq$fixed.acidity)

#summary of Volatile acidity
summary(wwq$volatile.acidity)
#The minimum and maximum are outliers.
#Median is 0.26 and mean is 0.2782


#Plotting the Volatile Acidity and using a binwidth of 0.1
qplot(volatile.acidity, 
      data = wwq,
      binwidth = 0.1,
      xlab = 'Volatile Acidity',
      color = I('black'),
      fill = I('pink')) +
      scale_x_continuous(limits = c(0,0.8))


#Plotting the citric acid content, binwidth of 0.1 
qplot(citric.acid, 
      data = wwq,
      binwidth = 0.1,
      xlab = 'Citric Acid',
      color = I('red'),
      fill = I('black')) +
      scale_x_continuous(limits = c(0, 1), breaks = c(0, 1, 0.1))

#The maximum is an outlier it is very far from the mean
summary(wwq$citric.acid)

#Plotting the residual sugar content 
qplot(residual.sugar, 
      data = wwq,
      xlab = 'Residual Sugar',
      color = I('black'),
      fill = I('pink')) +
  scale_x_continuous(limits = c(0,20), breaks = c(0, 5, 10, 15, 20))

#the maximum is an outlier and a bad data
summary(wwq$residual.sugar)


#plotting the chloride content and limiting the x-axis from 0 to 0.1
qplot(chlorides, 
      data = wwq,
      binwidth = 0.01,
      xlab = 'Chloride Content',
      color =I('black'),
      fill = I('pink'))+
      scale_x_continuous(limits = c(0, 0.1))

#The distribution of variables here is positively skewed with 75% of #the distribution falling below 0.05 and the 25% between 0.05 and 0.346
summary(wwq$chlorides)

#Plotting the Free Sulfur dioxide
qplot(free.sulfur.dioxide, 
      data = wwq,
      binwidth = 5) +
      scale_x_continuous(limits = c(0,110), breaks = seq(0,140,20))

#The distribution of variables is positively skewed with majority of
#the distribution under 75%. The mean is 35.31 and the median is 34.00
summary(wwq$free.sulfur.dioxide)


#Plotting the Total sulfur dioxide
qplot(total.sulfur.dioxide, data = wwq,
      binwidth = 10,
      xlab = 'Total Sulfur Oxide',
      color = I('black'),
      fill = I('red')) +
      scale_x_continuous(limits = c(0,300), breaks = seq(0, 300, 10))

#The variable distribution is positively skewed and the mean is 138 while the median is 134.
summary(wwq$total.sulfur.dioxide)

#Plotting the density
qplot(density, 
      data = wwq, 
      binwidth = 0.001,
      xlab = 'Density',
      color = I('black'),
      fill = I('red')) +
  scale_x_continuous(limits = c(0.98 ,1.02))

#the mean and the median are very close to each other, 0.994 and 0.9937
#respectively
summary(wwq$density)

#Plotting the pH 
qplot(pH, 
      data = wwq,
      xlab = 'pH',
      color = I('black'),
      fill = I('purple')) +
  scale_x_continuous(limits = c(2.7, 3.82), 
                     breaks = seq(2.7, 3.82, 0.1))

#The distribution of the variables are positively skewed.
#Most wines have a pH ranging from 2.5 to 3.9
summary(wwq$pH)

#Plotting Sulphates
qplot(sulphates, 
      data = wwq,
      xlab = 'Potassium Sulphates',
      color = I('black'),
      fill = I('blue')) +
  scale_x_continuous(limits = c(0.2, 1.08), 
                     breaks = seq(0.2, 1.08, 0.1))

#75% of the variable distribution is below 0.55
#The median is 0.47 and the mean is 0.4898
summary(wwq$sulphates)

#Plotting Alcohol
qplot(alcohol, 
      data = wwq, 
      binwidth = .25,
      xlab = 'Alcohol Content',
      color = I('black'),
      fill = I('dark grey')) +
  scale_x_continuous(limits = c(8, 14), 
                     breaks = seq(8, 14, 1))

#The median is 10.4 and the mean is 10.51. Variable distribution is #slightly positively skewed
summary(wwq$alcohol)


#Quality was measured by taking the average of the 3 expert ratings
#The experts rated most of wine with a quality of 6
#Only 5 white wines were rated a quality of 9 in our dataset
summary(wwq$quality)
#Plotting the quality
ggplot(aes(x = quality), data = wwq) + 
  geom_bar(fill = 'red') +
  xlab('White Wine Quality')





```

# Univariate Analysis

### What is the structure of your dataset?
My data set contains the 4898 observations and 11 numerical variables of chemical properties. I converted the X and quality variables to a factor variable.
```{r echo=FALSE}
str(wwq)
```

### What is/are the main feature(s) of interest in your dataset?

This data set contains 4,898 white wines with 11 variables on quantifying the chemical properties of each wine. At least 3 wine experts rated the quality of each wine, providing a rating between 0 (very bad) and 10 (very excellent).
Our Guiding Question is which chemical properties influence the quality of white wines.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
The variables fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, density, free sulfur dioxide, total sulfur dioxide, pH, sulphates, alcohol all contribute to the better quality of white wines.

```{r echo=FALSE}
#names of my dataset
names(wwq)
```


### Did you create any new variables from existing variables in the dataset?
Not really. I however passed my dataset into a new variable called wwq (white wine quality)

# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots}


#Creating a bucket for the quality. Rating wines from 0-4 as bad,
# 5-7 as average and 8-10 as very good
wwq$quality.bucket <- cut(as.numeric(wwq$quality), 
                          breaks = c(0, 4,5, 7, 9),
                          labels = c('poor', 
                                     'average',
                                     'better',
                                     'exquisite'))

#Plotting the alcohol variable against the quality bucket 
#that has been cut into 4 bins. We see that the white wines rated 
#between 4-5 have the least alcohol content. 
alcohol_quality <- ggplot(aes(x = quality.bucket, 
           y = alcohol), 
       data = wwq) + 
  geom_boxplot() + 
  coord_cartesian(ylim = c(9,13)) +
  stat_summary(fun.y = mean,
               geom = 'point',
               shape = 4)

names(wwq)

#measuring the correlation between chloride variable and quality 
cor.test(wwq$chlorides, as.numeric(wwq$quality))
#We have a correlation of -0.2099344 between the chloride variable
# and the quality variable
chlorides_quality <- ggplot(aes(x = quality.bucket, 
           y = chlorides), data = wwq) +     
  geom_boxplot() 

#testing the correlation between alcohol and the quality variable

cor.test(wwq$alcohol, as.numeric(wwq$quality))
#we observe a correlation of 0.4355747. A positive correlation
#indicating a strong relationship between the ratings generally
#by the wine experts.

#Plotting the boxplots of density and the quality buckets
density_quality <- ggplot(aes(x= quality.bucket, y = density), data = wwq) +
  geom_boxplot() +
  coord_cartesian( ylim = c(0.99,1))+
  stat_summary(fun.y = mean,
               geom = 'point',
               shape = 3)
library(gridExtra)
grid.arrange(alcohol_quality, density_quality, ncol = 1)  
#wwq %>%
 # group_by(quality)%>%
#summarise(mean_quality = mean(quality))
 # ungroup()
```


# Density Versus Alcohol Plot

```{r echo=FALSE}
#Plotting Desnity versus Alcohol
ggplot(aes(x = density, y = alcohol), 
       data = wwq) + 
  geom_jitter(alpha = 0.4)+
  coord_cartesian(xlim = c(0.98,1.01),
                  ylim = c(8,14))
  
```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

I checked the correlation between alcohol variable and quality variable
and  found both variables to be strongly positively correlated. The 
correlation between the two variables alcohol and quality is 
0.435547.
On checking the correlation between chloride variable and quality,
I found both variables to be weakly negatively correlated. The 
correlation between the two variables chlorides and quality is
-0.2099344

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
Another interesting relationship i observed is that, on plotting
the alcohol variable against density, we observe a negative slope.

### What was the strongest relationship you found?

The strongest relationship would be that alcohol content is a major 
factor that determines the quality of white wines. More investigations
need to be done on this.



# Multivariate Plots Section


###  Alcohol, Density, Quality

Alcohol has a strong correlation to quality of wine. Density and 
alcohol have a negative correlation of -0.7801376

```{r echo=FALSE}
#correlation test of alcohol with density
cor.test(wwq$density, wwq$alcohol)

```

# Alchol, Density and Quality cont'd
```{r echo=FALSE, Multivariate_Plots}

#visualizing the relationship between alcohol, density and quality
ggplot(aes(x = alcohol, y = density), 
       data = wwq) +
  geom_point(aes(color = quality.bucket))+
  coord_cartesian(ylim = c(0.985, 1.005)) +
  geom_vline(xintercept = 11, linetype = 2) +
  geom_hline(yintercept = 0.995, linetype = 2)
#Wines rated poor have a low alcohol content.


```

# Alcohol, Residual Sugar and Density

```{r echo=FALSE}
#visualizing Residual Sugar and density
ggplot(aes(x = residual.sugar,
           y = density), data = wwq) +
  geom_point(aes(color = alcohol ), alpha = 1/5)+
  coord_cartesian(xlim = c(0,30), 
                  ylim = c(0.985, 1.005)) +
  geom_smooth()
  
#Visualizing Alcohol, Residual sugar, Density, and quality
#Dealing only with the ratios
wwq <- transform(wwq, 
                 sugar.alcohol.ratio = residual.sugar/alcohol)
ggplot(aes(x= sugar.alcohol.ratio,
           y = density), data = wwq) +
  geom_point(aes(color = quality.bucket) ) +#, alpha = 1/10)+
  coord_cartesian(xlim = c(0,3), ylim = c(0.985, 1.005))+
  geom_smooth()
#The smooth line shows that wines rated average have a lower
#density than wines rated poor

#Plotting residual sugar and alcohol
ggplot(aes(x = log10(residual.sugar), y = alcohol), data = wwq) +
  geom_jitter(aes(color = quality.bucket))+
  coord_cartesian(xlim = c(0, 2), ylim = c(8, 14)) 

```


# Multivariate Analysis


### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?
### Were there any interesting or surprising interactions between features?
From our first plot of density against alcohol we can observe that 
on wines with higher densities are rated poor while wines 
with lower densities are rated better in term of quality.
It is also observed that wines rated poor have a lower alcohol
content. Perhaps, our experts think that the more the alcohol
in the wine, the better the wine served.


------

# Final Plots and Summary

### Plot One
# Alcohol and Quality Bucket.

```{r echo=FALSE, Plot_One}
#visualizing Alcohol and quality bucket relationship
ggplot(aes(x = quality.bucket,
           y = alcohol), data = wwq) +
  geom_boxplot() +
  stat_summary(fun.y = mean, 
               geom = 'point',
               shape = 4) +
  coord_cartesian(ylim = c(9,13))+
  labs(title = 'Alcohol by Quality')+
  ylab('Alcohol')+
  xlab('Quality')

```

### Description One
Average and Better rated wines have a alcohol content of 11 - 13.
Poor wines have lower alcohol content as rated by the experts.


### Plot Two
# Residual Sugar and Density
```{r echo=FALSE, Plot_Two}
#visualizing relationship between residual sugar and density
ggplot(aes(x = log10(residual.sugar),
           y = density), data = wwq) +
  geom_jitter(aes(color = alcohol ))+
  coord_cartesian(xlim = c(0,1.5), 
                  ylim = c(0.985, 1.005)) +
  geom_smooth(method = 'auto', color = 'red')+
  labs(title = 'Residual Sugar vs Density')+
  xlab('Residual Sugar') +
  ylab('Density')

```

### Description Two
Better rated wines have a lower density than poorly rated wines.

### Plot Three

# Residual Sugar and Alcohol
```{r echo=FALSE, Plot_Three}

# Visualizing Residual sugar and Alcohol
ggplot(aes(x = log10(residual.sugar), 
           y = alcohol), data = wwq) +
  geom_jitter(aes(color = quality.bucket))+
  coord_cartesian(xlim = c(0, 1.5), 
                  ylim = c(8, 14)) +
  geom_smooth() +
  labs(title = 'Alcohol vs Residual Sugar')+
  xlab('Residual Sugar')+
  ylab('Alcohol')

```

### Description Three

Average and better rated wines have higher residual sugar.
```{r echo=FALSE}
#correlation between residual sugar and alcohol

cor.test(wwq$residual.sugar, wwq$alcohol)
#correlation between residual sugar and alcohol being tested
# as 0.4506312
```

------

# Reflection

The white wine dataset contains 11 distinct chemical properties of 4898
samples of wine as rated by 3 experts. Each of the wine sampled were 
rated on the quality of the wine.
Our goal was to find which properties would most affect the quality of the wine. Checking the correlation between alcohol and the quality of
wine, I observed that there was a strong positive correlation between 
the two variables.
I visualized and the relationship between each variables with each
other to see which would likely affect the quality of the wine rating

